---
title: "Connecticut Wild Edibles Phenology"
author: "Cody Limber"
date: today
format: html
---

## Goal

Determine the best weeks of the year to observe different wild edible
species in Connecticut by analyzing iNaturalist observation data.
Produce a season-period calendar showing when each species is active
and at peak abundance.

## Setup

```{r setup}
#| message: false
library(tidyverse)
library(rinat)
library(rgbif)
library(lubridate)
library(plotly)
library(DT)
library(sf)
library(tigris)
library(jsonlite)
library(yaml)

source("R/phenology_engine.R")

params <- load_params("params.yml")

out_dir <- "outs"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
cache_dir <- file.path(out_dir, "cache")
```

## Fetch observations

The engine paginates by year and month to stay within API limits (10k for iNat,
100k for GBIF).

```{r fetch}
#| cache: true
obs_raw <- fetch_observations(params, cache_dir = cache_dir)
cat("Total", params$quality, "-grade observations:", nrow(obs_raw), "\n")
```

## Clean and summarize

```{r clean}
cleaned <- clean_observations(obs_raw, min_obs = params$min_obs)
obs_filtered    <- cleaned$obs
species_counts  <- cleaned$species_counts
cat(nrow(species_counts), "species with >=", params$min_obs, "observations\n")
```

## Assign observations to counties

```{r county-assignment}
#| cache: true
county_data <- assign_counties(obs_filtered, "CT",
                                params$county_year)
cat("Observations with coordinates:", nrow(county_data$obs_county), "\n")
cat("Location-hidden:", county_data$location_hidden_count, "\n")
cat("County distribution:\n")
count(county_data$obs_county, county, sort = TRUE)
```

## Rarity tiers and conservation status

```{r rarity}
species_counts <- assign_rarity(species_counts,
                                 params$rarity$rare_below,
                                 params$rarity$uncommon_below)
cat("Rarity distribution:\n")
count(species_counts, rarity)
```

```{r conservation-status}
#| cache: true
# Build place pattern for NatureServe lookup
place_pattern <- "Connecticut|CT"
if (place_pattern == "") place_pattern <- NULL

cons_data <- fetch_conservation_all(obs_filtered, place_pattern,
                                    data_source = params$data_source)

species_counts <- species_counts |>
  select(-any_of(c("taxon_id", "conservation", "concern_level"))) |>
  left_join(cons_data$taxon_ids, by = "species") |>
  left_join(cons_data$conservation, by = "taxon_id")

cat("\nConservation status distribution:\n")
count(species_counts, concern_level)
```

## Compute weekly phenology

```{r phenology}
pheno <- compute_phenology(obs_filtered, species_counts, params)
weekly          <- pheno$weekly
species_summary <- pheno$species_summary

multi <- species_summary |> filter(period_count > 1)
if (nrow(multi) > 0) {
  cat("Multi-", params$terminology$multi_period, " species:\n", sep = "")
  print(multi |> select(common_name, species, period_count))
}
```

## Season period calendar

```{r calendar-plot}
#| fig-width: 5
#| fig-height: 20
cal <- build_calendar_plot(weekly, species_summary, params)

# Show interactive version (falls back to static if plotly fails)
if (!is.null(cal$plotly)) cal$plotly else cal$gg

# Save static version
ggsave(file.path(out_dir, "wild_edible_calendar.png"),
       plot = cal$gg, width = 10, height = 18, dpi = 300)
```

## Species summary table

```{r summary-table}
summary_out <- build_summary_table(species_summary, params)
summary_out$dt

write_csv(summary_out$table,
          file.path(out_dir, "species_season_periods.csv"))
```

## What's available this week?

```{r whats-active}
active_now <- whats_active(Sys.Date(), weekly, species_summary, params)
datatable(active_now,
          options = list(pageLength = 25, dom = "ft"),
          caption = paste("What's available the week of",
                          format(Sys.Date(), "%B %d, %Y")))
```

```{r whats-active-monthly}
active_months <- if (!is.null(params$active_months)) {
  params$active_months[1]:params$active_months[2]
} else {
  4:10
}

monthly_active <- build_monthly_overview(weekly, species_summary, params,
                                          months = active_months)
datatable(monthly_active,
          filter = "top",
          options = list(pageLength = 25, autoWidth = TRUE),
          caption = paste("Monthly", tolower("Wild Edible"),
                          "activity overview"))
```

## Export JSON

```{r json-export}
# Export to outs/ for archival and to repo root for the website
export_json(weekly, species_summary, params, obs_raw,
            county_data = county_data,
            out_path = file.path(out_dir, "phenology_data.json"))
file.copy(file.path(out_dir, "phenology_data.json"), "phenology_data.json",
          overwrite = TRUE)
cat("Copied phenology_data.json to repo root for website\n")
```

## Build info

```{r build-info}
write_build_info(out_dir, params, obs_raw, nrow(species_counts))
```
